Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-22 analysis/paper/paper.tex
reduce code in paper

Word count: 4722

@@

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues (1991) and
Shellis (1978). In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins.

@@

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor
with no history of caries, who abstained from oral hygiene for 24 hours, and no
food was consumed two hours prior to donation. No antibiotics were taken up to six
months prior to donation.
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15\% active chlorine)
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised
20\% (v/v) glycerine for four hours at 36°C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the
substrata were transferred to a new plate containing 1 ml/well AS and incubated
in a shaking incubator (Infors HT Ecotron) at 36°C, 30 rpm.
The inoculation process was repeated on days 3 and 5.
AS was partially refreshed once per day and fully refreshed every three days,
throughout the experiment, by transferring the substrata to a new plate containing
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing
5\% (w/v) sucrose, for six minutes twice daily, except on inoculation days
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l Urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see \ref{fig:protocol-fig} for an overview of the protocol).

All laboratory work was conducted in sterile conditions under a laminar flow hood
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment, were included to detect starch contamination from the
environment, or cross-contamination from other wells in the same plate.

@@

A 1 ml aliquot of each starch solution was taken, from which 10 \(\mu\)l was mounted
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts
were extrapolated (see Supplementary material for more details). For potato
treatment samples, the whole slide was counted.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. Wheat and mix counts were extrapolated as described above.
Both wheat and potato granules were divided into three size categories:
small (\textless10 \(\mu\)m), medium (10 -- 20 \(\mu\)m), and large (\textgreater20 \(\mu\)m).

@@

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg
calculus (also z-score standardised) to account for differences in starch
concentration in the calculus (as per Wesolowski et al., 2010).

@@

It was not possible to differentiate between potato and
wheat starches smaller than ca. 10 \(\mu\)m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. This is a
reasonable assumption to make based on the small granule counts from the separate
treatment solutions, where the wheat granules made up
99.2\%
of all small granule counts in the separated solutions.

@@

\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.}
\centering
\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
\hline
Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
\hline
mix & potato &  &  & 1959 (79.6\%) & 1800 & 501 (20.40\%) & 446 & 2460 (100\%) & 2190\\
\hline
mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6030 & 1381 (7.93\%) & 1200 & 17417 (100\%) & 15900\\
\hline
mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7650 & 1882 (9.47\%) & 1600 & 19877 (100\%) & 17800\\
\hline
potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2400 & 930 (19.20\%) & 929 & 4846 (100\%) & 3320\\
\hline
wheat & wheat & 15235 (55.00\%) & 11900 & 12148 (43.9\%) & 11100 & 1953 (7.06\%) & 2020 & 27680 (100\%) & 23600\\
\hline
\end{tabular}
\end{table}

The separate wheat and potato solutions were made with a 0.25\% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25\% (w/v) of each
starch, with a total concentration of 0.50\% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = \ensuremath{1.98\times 10^{6}}, \ensuremath{2.7\times 10^{7}}),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules
(mean = \ensuremath{2.77\times 10^{4}}).
The potato treatment had the lowest absolute counts in both the solution
(\ensuremath{3.02\times 10^{6}})
and in the biofilm samples
(4850)
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).

@@

\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
              were incorporated into the samples.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Treatment & Starch & Small & Medium & Large & Total\\
\hline
mix & potato &  & 0.186\% & 0.054\% & 0.124\%\\
\hline
mix & wheat & 0.051\% & 0.102\% & 0.077\% & 0.064\%\\
\hline
mix & both & 0.051\% & 0.114\% & 0.069\% & 0.069\%\\
\hline
potato & potato & 0.284\% & 0.266\% & 0.060\% & 0.161\%\\
\hline
wheat & wheat & 0.094\% & 0.189\% & 0.069\% & 0.109\%\\
\hline
\end{tabular}
\end{table}

The proportion of total starches from the solutions that were incorporated
into the samples ranged from
0.064\% to 0.161\%,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \ref{tab:sample-prop-tbl}).
There is an inverse relationship between the absolute starch count in the solutions
and the proportional incorporation of starches in the biofilm samples, i.e., potato
had the lowest absolute count in solutions, but the highest proportional
incorporation, and vice versa for the mixed treatment.

Wheat incorporation was most affected in the mixed-treatment samples, with only
0.064\% of the total available starches being incorporated into
the sample, compared to
0.161\%
in the separated wheat treatment.

@@

{\centering \includegraphics[width=0.8\linewidth]{../figures/ratio-plots-1}

@@

Overall, medium starch granules had a higher mean rate of incorporation
(0.186\%)
than small
(0.143\%)
and large
(0.065\%)
starch granules across all treatments, while large potato starches had the lowest
rate of incorporation across all treatments.

@@

{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot-1}

@@

\caption{Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}\label{fig:cor-plot}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot2-1}

@@

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.064\% to 0.161\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.064\% to 0.161\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010), which could in part be due to the
lack of oral amylase activity in our model.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
diameter were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot});
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot2}).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2019; Radini et al., 2017).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calulus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface of the dental calculus, as well as unmineralised islands
and channels may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
Future study should explore how burial affects the recovery of starch from the
biofilm model.

A limitation of this study
The absence of \(\alpha\)-amylase in the model is a limitation of this study, as
the total granule counts were not subject to hydrolysis. This would likely have
reduced and affected the size ratios, as smaller starches may be more
susceptible to hydrolysis
(Franco et al., 1992; Haslam, 2004); however,
the absence can also allow us to directly explore
the effect of \(\alpha\)-amylase on starch counts in future experiments,
where \(\alpha\)-amylase can be added to the model
in concentrations similar to those found in the oral cavity (Scannapieco et al., 1993).
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and
wheat solutions of the same concentration (w/v). Further research should test
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown
negligible morphological changes caused by exposure to EDTA
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that
was feasible to count in a reasonable time period, so we calculated the total
counts by extrapolating from three slide transects.
Thus, we reasonably assume that the three transects are a good representation
of the entire slide, and that the distribution of all granules on the slide is
relatively homogeneous.\\
Finally, we only used native starches in the experimental procedure and the results
will likely differ for processed starches (García-Granero, 2020).
Based on the comparatively low counts obtained by Leonard
and colleagues (2015, Supplement 2), processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

>>>>

@@

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues (1991) and
Shellis (1978). In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins. The individual pegs
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions,
and mineral solutions.

@@

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor
with no history of caries, who abstained from oral hygiene for 24 hours. No
food was consumed two hours prior to donation and no antibiotics were taken up
to six months prior to donation.
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15\% active chlorine)
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised
20\% (v/v) glycerine for four hours at 36°C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the
substrata were transferred to a new plate containing 1 ml/well AS and incubated
in a shaking incubator (Infors HT Ecotron) at 36°C, 30 rpm.
The inoculation process was repeated on days 3 and 5.
AS was partially refreshed once per day and fully refreshed every three days,
throughout the experiment, by transferring the substrata to a new plate containing
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing
5\% (w/v) sucrose, for six minutes twice daily, except on inoculation days
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see \ref{fig:protocol-fig} for an overview of the protocol).

All laboratory work was conducted in sterile conditions under a laminar flow hood
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment were included to detect starch contamination from the
environment or cross-contamination from other wells in the same plate.

@@

A 1 ml aliquot of each starch solution was taken, from which 10 \(\mu\)l was mounted
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts
were extrapolated to the total number of granules exposed to the samples over 16
days of treatments (see Supplementary Material for more details). For potato
treatment samples, the whole slide was counted.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material).

Both wheat and potato granules were divided into three size categories:
small (\textless10 \(\mu\)m), medium (10 -- 20 \(\mu\)m), and large (\textgreater20 \(\mu\)m).

@@

Pearson's \emph{r} was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg
calculus (also z-score standardised) to account for differences in starch
concentration in the calculus (as per Wesolowski et al., 2010).

@@

It was not possible to differentiate between potato and
wheat starches smaller than ca. 10 \(\mu\)m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the total amount of small starches in these two solutions,
0.653\%
are from wheat.

@@

\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.}
\centering
\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
\hline
Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
\hline
mix & potato &  &  & 1959 (79.6\%) & 1801 & 501 (20.40\%) & 446 & 2460 (100\%) & 2189\\
\hline
mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6026 & 1381 (7.93\%) & 1196 & 17417 (100\%) & 15878\\
\hline
mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 4929 & 1882 (9.47\%) & 989 & 19877 (100\%) & 13451\\
\hline
potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2402 & 930 (19.20\%) & 929 & 4846 (100\%) & 3316\\
\hline
wheat & wheat & 15235 (55.00\%) & 11944 & 12148 (43.9\%) & 11052 & 1953 (7.06\%) & 2016 & 27680 (100\%) & 23554\\
\hline
\end{tabular}
\end{table}

The separate wheat and potato solutions were made with a 0.25\% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25\% (w/v) of each
starch, with a total concentration of 0.50\% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = \ensuremath{2.9\times 10^{7}}),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules
(mean = \ensuremath{2.77\times 10^{4}}).
The potato treatment had the lowest absolute counts in both the solution
(\ensuremath{3.02\times 10^{6}})
and in the biofilm samples
(4850)
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).

@@

\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
              were incorporated into the samples.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Treatment & Starch & Small & Medium & Large & Total\\
\hline
mix & potato &  & 0.19\% & 0.05\% & 0.12\%\\
\hline
mix & wheat & 0.05\% & 0.10\% & 0.08\% & 0.06\%\\
\hline
mix & both & 0.05\% & 0.11\% & 0.07\% & 0.07\%\\
\hline
potato & potato & 0.28\% & 0.27\% & 0.06\% & 0.16\%\\
\hline
wheat & wheat & 0.09\% & 0.19\% & 0.07\% & 0.12\%\\
\hline
\end{tabular}
\end{table}

The proportion of total starches from the solutions that were incorporated
into the samples ranged from
0.06\% to 0.16\%,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \ref{tab:sample-prop-tbl}).
There is an inverse relationship between the absolute starch count in the solutions
and the proportional incorporation of starches in the biofilm samples, i.e., potato
had the lowest absolute count in solutions, but the highest proportional
incorporation, and vice versa for the mixed treatment.

Wheat incorporation was most affected in the mixed-treatment samples, with only
0.06\% of the total available starches being incorporated into
the sample, compared to
0.16\%
in the separated wheat treatment.

@@

{\centering \includegraphics{../figures/ratio-plots-1}

@@

Overall, medium starch granules had a higher mean rate of incorporation
(0.171\%)
than small
(0.120\%)
and large
(0.066\%)
starch granules across all treatments, while large potato starches had the lowest
rate of incorporation across all treatments.

@@

{\centering \includegraphics{../figures/cor-plot-1}

@@

\caption{Scatter plot of sample weight and standardised starch count by z-score for separated treatments.}\label{fig:cor-plot}
\end{figure}

@@

{\centering \includegraphics{../figures/cor-plot2-1}

@@

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.06\% to 0.16\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.06\% to 0.16\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010), which could in part be due to the
lack of oral amylase activity in our model.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot});
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot2}).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2019; Radini et al., 2017).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calculus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface of the dental calculus, as well as unmineralised islands
and channels may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
Future study should explore how burial affects the recovery of starch from the
biofilm model.

The absence of \(\alpha\)-amylase in the model is a limitation of this study, as
the total granule counts were not subject to hydrolysis. This would likely have
reduced and affected the size ratios, as smaller starches may be more
susceptible to hydrolysis
(Franco et al., 1992; Haslam, 2004); however,
the absence can also allow us to directly explore
the effect of \(\alpha\)-amylase on starch counts in future experiments,
where \(\alpha\)-amylase can be added to the model
in concentrations similar to those found in the oral cavity (Scannapieco et al., 1993).
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and
wheat solutions of the same concentration (w/v). Further research should test
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown
negligible morphological changes caused by exposure to EDTA
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that
was feasible to count in a reasonable time period, so we calculated the total
counts by extrapolating from three slide transects.
Thus, we reasonably assume that the three transects are a good representation
of the entire slide, and that the distribution of all granules on the slide is
relatively homogeneous.\\
Finally, we only used native starches in the experimental procedure and the results
will likely differ for processed starches (García-Granero, 2020).
Based on the comparatively low counts obtained by Leonard
and colleagues (2015, Supplement 2), processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.
